---
title: "Final report analysis"
author: "Borna Bejuk, Josip Silović"
date: "January 28, 2018"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
library(tidyverse)
library(stringr)
library(ggplot2)
knitr::opts_chunk$set(results = 'hold')
source("../R/data_reading.R")
options(warn = 0)
```


<h1> Game of Thrones deaths anaylsis </h1>

 <h3>
 1. Data summary:
 </h3>
 
```{r echo=FALSE}
got_deaths <- tbl_df(got_deaths)
summary(got_deaths)
cat("--------------------------------------------------------------------------------------\n")
str(got_deaths)
cat("--------------------------------------------------------------------------------------\n")
head(got_deaths)
```
 
```{r, eval = F}
VARIABLE DESCRIPTIONS:
Name                Character name and surname

Allegiances         Character house allegiance

Death Year          Year of characters death

Book of Death       Book number of characters death

Death  Chapter      Chapter of characters death

Book Intro Chapter  Chapter character was introduced in

Gender              Character gender
                    (0 = Female, 1 = Male)

Nobility            Nobility status
                    (0 = Commoner, 1 = Noble)

GoT                 Appeared in first book
                    (0 = No, 1 = Yes)

CoK                 Appeared in second book
                    (0 = No, 1 = Yes)

SoS                 Appeared in third book
                    (0 = No, 1 = Yes)

FfC                 Appeared in fourth book
                    (0 = No, 1 = Yes)

DwD                 Appeared in fifth book
                    (0 = No, 1 = Yes)
```
 
<h3>
2. Lets get the answer to the basic question - the distribution of deaths per book.
</h3>

```{r echo=FALSE}
options(warn = 0)
graph <- ggplot(got_deaths, aes(x = Book.of.Death, fill = as.factor(Allegiances)))

graph <- graph + geom_histogram(bins = 5, color = "black", na.rm = TRUE) + labs(fill = "House Allegiances", x = "Book of death", y = "Number of deaths", title = "Number of deaths each book with allegiance distinctions")

graph
#ggsave(plot = graph, filename="../figures/books_deaths_by_allegiances.png")
```

- Unfortunately, there is a lot of NA values inside the dataset, and there is no way of filling that data given the other variables, so the first graph shows us only about 300 deaths out of total of 917 character deaths in the dataset
- The naive assumption would be to fill the book of death with the last book they appeared in, however than can only mean they were mentioned in that book, not that they were actually alive at that time
- Some first glance observation shows us that having no Allegiance was not a good idea in any of the books so far

<h3>
3. Which house has suffered the most deaths in the series?
</h3>

```{r echo=FALSE}
options(warn = -1)
graph <- ggplot(got_deaths, aes(x = Allegiances))

graph <- graph + geom_histogram(fill = "light blue", color = "black", na.rm = TRUE, stat = "count") + theme(axis.text.x  = element_text(angle = 90)) + labs(x = "Allegiances", y = "Number of allegiant", title = "Number of allegiants for each House")

graph
#ggsave(plot = graph, filename="../figures/allegiances_deaths_histogram.png")
```


- Looks like our first glance intuition was correct, neutral was definitely not a good option
- Other than neutral fraction deaths, the top contenders for most deaths are: Night's Watch, Lannisters and Starks being almost equa 
. Night's Watch was to be expected, given the fact that they are the first line of defense against the Wildlings and the Army of the Dead

<h3>
4. Lets observe the deaths of Night's Watch, Lannisters and Starks a little closer.
</h3>

```{r echo=FALSE}
options(warn = -1)
most_deaths <- filter(got_deaths, Allegiances == c("Night's Watch", "Lannister", "Stark"))

graph <- ggplot(most_deaths, aes(x = Book.of.Death, fill = as.factor(Allegiances)))

graph <- graph + geom_histogram(bins = 5, color = "black", na.rm = TRUE, binwidth = 0.25, position = "dodge") + labs(fill = "House Allegiances", x = "Book of death", y = "Number of deaths", title = "Number of deaths each book for Night's Watch, Lannister and Stark Houses")

graph
#ggsave(plot = graph, filename="../figures/top_houses_deaths_hist.png")
```


- Given the histogram above, it seems that after the first killings of Starks in Book 1, the next 2 seasons gave rise to series of vengeances, resulting in deaths in both Starks and Lannisters Houses. Poor members of Night's Watch seem to be very active in all seasons...

<h3>
5. Okay, we all know that the whole series is filled with bloodbath, but does having a Nobility status make any difference, or is the status not important?
</h3>

```{r echo=FALSE}
options(warn = -1)
graph <- ggplot(got_deaths, aes(x = Book.of.Death, fill = as.factor(Nobility)))

graph <- graph + geom_histogram(bins = 5, color = "black", na.rm = TRUE, binwidth = 0.5, position = "dodge") + labs(fill = "Nobility status", x = "Book of death", y = "Number of deaths", title = "Number of deaths each book with nobility distinction")

graph
#ggsave(plot = graph, filename="../figures/nobility_status_deaths_hist.png")
```

<h3>
6. Last but not least, lets see the if George R.R. Martin has a soft spot for ladies, or is his pen simply merciless.
</h3>

```{r echo=FALSE}
options(warn = -1)
graph <- ggplot(got_deaths, aes(x = Book.of.Death, fill = as.factor(Gender)))

graph <- graph + geom_histogram(bins = 5, color = "black", na.rm = TRUE, binwidth = 0.5, position = "dodge") + labs(fill = "Gender", x = "Book of death", y = "Number of deaths", title = "Number of deaths each book with gender distinction (0 - Female, 1 - Male)")

graph
#ggsave(plot = graph, filename="../figures/gender_status_deaths_hist.png")
```

<h1> Highest Mountains analysis </h1>


## Data summary
```{r, echo=FALSE}
hm <- read.csv("../data/HighestMountains.csv", stringsAsFactors = F, header = T, na.strings = '')
hm <- tbl_df(hm)
head(hm)
glimpse(hm)
summary(hm)
```
```{r, echo = F}
hm <- mutate(hm, HeightCategory = cut(Height..m., 5, labels = F))
hm <- mutate(hm, RankCategory = ntile(Rank, 5))
hm$First.ascent <- as.integer(sub('unclimbed', NaN, hm$First.ascent))
hm <- mutate(hm, FirstClimbCategory = cut(First.ascent, 5, dig.lab = 4, na.rm = T))
```
## 1. Let us first see the distribution of mountain heights in 10 bins
```{r, echo = F}
options(warn = -1)
graph <- ggplot(hm, aes(x = Height..m.)) + geom_histogram(bins = 10, fill = 'light blue', color = 'black') + labs(x = 'Mountain height in meters', y = 'Count', title = 'Distribution of mountain heights')
#ggsave(plot = graph, filename="../figures/MountainHeights.png")
graph
```

- The distribution is right skewed (median is to the left of the mean), meaning we have a small number of mountains that fall to the right of the mean and a large number that fall to the left of the mean.  

## 2. Grouping the mountain heights into 5 groups based on their height
```{r, echo = F}
graph <- ggplot(hm, aes(x = HeightCategory)) + geom_histogram(bins = 5, stat = 'count', fill = 'light blue', color = 'black') + labs(x = 'Height category', y = 'Count', title = 'Height count when height is categorized into 5 groups')
#ggsave(plot = graph, filename="../figures/MountainHeights5groups.png")
graph
```

- We decided to divide the mountains into 5 categories based on their height to aid us in further analysis of the dataset
- Based on this initial grouping we see the most mountains fall into the first category

## 3. Let us see who are the parent mountains of our mountains when looking at our 5 groups
```{r, echo = F}
filter(hm, !is.na(hm$Parent.mountain)) %>% ggplot(aes(x = HeightCategory, fill = Parent.mountain)) + geom_histogram(bins = 5, color = 'black') + labs(fill = 'Parent mountain', x = 'Height category', y = 'Count', title = 'Height count based on the mountain\'s parent mountain\n\n') -> graph
#ggsave(plot = graph, filename="../figures/MountainHeightsByParent.png")
graph
```

- We wanted the see if we could spot any link between the mountain's height and its parent mountain
- We can see that the highest mountains mostly have Mount Everest as its parent mountain

## 4. Is there any link between the mountain height and the number of failed attempts ?
```{r, echo = F}
graph <- ggplot(hm, aes(x = Height..m., y = Failed.attempts.bef..2004, color = Height..m.)) + geom_point(na.rm = TRUE) + labs(x = 'Mountain height in meters', y = 'Number of failed ascent attempts', title = 'Number of failed ascent attempts in connection with the mountain height', color = 'Height')
#ggsave(plot = graph, filename="../figures/FailedAscentAttemptsPoints.png")
graph
```

- Due to small number of data points it is hard to draw any conclusions but a sort of exponential growth is noticable

We can try to visualize the pattern using stat_smooth

```{r, echo = F}
graph + geom_point(na.rm = TRUE) + stat_smooth(na.rm = TRUE, method = 'loess') -> graph2
#ggsave(plot = graph2, filename="../figures/FailedAscentAttemptsPointsLoess.png")
graph2
```

Or we could furhter try to improve our visualization by using different colors for each height category.
```{r, echo = F}
graph <- ggplot(hm, aes(x = hm$Height..m., y = hm$Failed.attempts.bef..2004, color = HeightCategory)) + geom_point(na.rm = TRUE) + geom_smooth(na.rm = TRUE, aes(x = hm$Height..m., y = hm$Failed.attempts.bef..2004), method = 'loess', linetype = 4, color = "grey", se = FALSE, inherit.aes = FALSE) + labs(x = 'Mountain height in meters', y = 'Number of failed ascent attempts', title = 'Number of failed ascent attempts in connection with the mountain height')
#ggsave(plot = graph, filename="../figures/FailedAscentAttemptsPointsLoessColor.png")
graph
```


## 5. Let us again see the number of failed attempts for each height category using histogram
```{r, echo = F}
options(warn = -1)
hm %>% group_by(HeightCategory) -> groupedByHC
summarise(groupedByHC, countOfFailures = sum(Failed.attempts.bef..2004, na.rm = T)) -> groupedByHC
graph <- ggplot(groupedByHC, aes(groupedByHC$HeightCategory, groupedByHC$countOfFailures)) + geom_col(fill = 'light blue', color = 'black') + labs(x = 'Height category', y = 'Failed attempts', title = 'Number of failed attempts for each height category')
#ggsave(plot = graph, filename="../figures/FailedAscentAttempts.png")
graph
```

- Using histogram we can see that the third category of mountains was the most 'trialed and errored'

## 6. Let's investigate in which years were the mountain hikers most active
```{r, echo = F}

filter(hm, !is.na(hm$FirstClimbCategory)) %>% ggplot(aes(FirstClimbCategory)) + geom_histogram(bins = 5, stat = 'count', na.rm = TRUE, fill = 'light blue', color = 'black') + labs(x = 'Height category', y = 'Count', title = 'Count of ascents when categorized into 5 groups by the year of ascent') -> graph
#ggsave(plot = graph, filename="../figures/YearOfAscents.png")
graph
```

- We see that according to this dataset most ascents fall in the category that is between years 1946 and 1962.



