---
title: "Game of Thrones character deaths analysis"
author: "Josip Silović"
date: "25/01/2018"
output:
  html_document: default
---

```{r setup, include=FALSE}
library(tidyverse)
library(stringr)
library(ggplot2)
knitr::opts_chunk$set(results = 'hold')
source("../R/data_reading.R")
options(warn = 0)
```


 <h3>
 1. Data summary:
 </h3>
 
```{r echo=FALSE}
got_deaths <- tbl_df(got_deaths)
summary(got_deaths)
cat("--------------------------------------------------------------------------------------\n")
str(got_deaths)
cat("--------------------------------------------------------------------------------------\n")
head(got_deaths)
```
 
```{r, eval = F}
VARIABLE DESCRIPTIONS:
Name                Character name and surname

Allegiances         Character house allegiance

Death Year          Year of characters death

Book of Death       Book number of characters death

Death  Chapter      Chapter of characters death

Book Intro Chapter  Chapter character was introduced in

Gender              Character gender
                    (0 = Female, 1 = Male)

Nobility            Nobility status
                    (0 = Commoner, 1 = Noble)

GoT                 Appeared in first book
                    (0 = No, 1 = Yes)

CoK                 Appeared in second book
                    (0 = No, 1 = Yes)

SoS                 Appeared in third book
                    (0 = No, 1 = Yes)

FfC                 Appeared in fourth book
                    (0 = No, 1 = Yes)

DwD                 Appeared in fifth book
                    (0 = No, 1 = Yes)
```
 
<h3>
2. Lets get the answer to the basic question - the distribution of deaths per book.
</h3>

```{r echo=FALSE}
options(warn = 0)
graph <- ggplot(got_deaths, aes(x = Book.of.Death, fill = as.factor(Allegiances)))

graph <- graph + geom_histogram(bins = 5, color = "black", na.rm = TRUE) + labs(fill = "House Allegiances", x = "Book of death", y = "Number of deaths", title = "Number of deaths each book with allegiance distinctions")

graph
#ggsave(plot = graph, filename="../figures/books_deaths_by_allegiances.png")
```

- Unfortunately, there is a lot of NA values inside the dataset, and there is no way of filling that data given the other variables, so the first graph shows us only about 300 deaths out of total of 917 character deaths in the dataset
- The naive assumption would be to fill the book of death with the last book they appeared in, however than can only mean they were mentioned in that book, not that they were actually alive at that time
- Some first glance observation shows us that having no Allegiance was not a good idea in any of the books so far

<h3>
3. Which house has suffered the most deaths in the series?
</h3>

```{r echo=FALSE}
options(warn = -1)
graph <- ggplot(got_deaths, aes(x = Allegiances))

graph <- graph + geom_histogram(fill = "light blue", color = "black", na.rm = TRUE, stat = "count") + theme(axis.text.x  = element_text(angle = 90)) + labs(x = "Allegiances", y = "Number of allegiant", title = "Number of allegiants for each House")

graph
#ggsave(plot = graph, filename="../figures/allegiances_deaths_histogram.png")
```


- Looks like our first glance intuition was correct, neutral was definitely not a good option
- Other than neutral fraction deaths, the top contenders for most deaths are: Night's Watch, Lannisters and Starks being almost equa 
. Night's Watch was to be expected, given the fact that they are the first line of defense against the Wildlings and the Army of the Dead

<h3>
4. Lets observe the deaths of Night's Watch, Lannisters and Starks a little closer.
</h3>

```{r echo=FALSE}
options(warn = -1)
most_deaths <- filter(got_deaths, Allegiances == c("Night's Watch", "Lannister", "Stark"))

graph <- ggplot(most_deaths, aes(x = Book.of.Death, fill = as.factor(Allegiances)))

graph <- graph + geom_histogram(bins = 5, color = "black", na.rm = TRUE, binwidth = 0.25, position = "dodge") + labs(fill = "House Allegiances", x = "Book of death", y = "Number of deaths", title = "Number of deaths each book for Night's Watch, Lannister and Stark Houses")

graph
#ggsave(plot = graph, filename="../figures/top_houses_deaths_hist.png")
```


- Given the histogram above, it seems that after the first killings of Starks in Book 1, the next 2 seasons gave rise to series of vengeances, resulting in deaths in both Starks and Lannisters Houses. Poor members of Night's Watch seem to be very active in all seasons...

<h3>
5. Okay, we all know that the whole series is filled with bloodbath, but does having a Nobility status make any difference, or is the status not important?
</h3>

```{r echo=FALSE}
options(warn = -1)
graph <- ggplot(got_deaths, aes(x = Book.of.Death, fill = as.factor(Nobility)))

graph <- graph + geom_histogram(bins = 5, color = "black", na.rm = TRUE, binwidth = 0.5, position = "dodge") + labs(fill = "Nobility status", x = "Book of death", y = "Number of deaths", title = "Number of deaths each book with nobility distinction")

graph
#ggsave(plot = graph, filename="../figures/nobility_status_deaths_hist.png")
```

