---
title: "Highest mountains exploratory analysis"
author: "Borna Bejuk"
date: "January 26, 2018"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(stringr)
library(ggplot2)
options(warn = 0)
```

## Data summary
```{r, echo=FALSE}
hm <- read.csv("../data/HighestMountains.csv", stringsAsFactors = F, header = T, na.strings = '')
hm <- tbl_df(hm)
head(hm)
glimpse(hm)
summary(hm)
```
```{r, echo = F}
hm <- mutate(hm, HeightCategory = cut(Height..m., 5, labels = F))
hm <- mutate(hm, RankCategory = ntile(Rank, 5))
hm$First.ascent <- as.integer(sub('unclimbed', NaN, hm$First.ascent))
hm <- mutate(hm, FirstClimbCategory = cut(First.ascent, 5, dig.lab = 4, na.rm = T))
```
## 1. Let us first see the distribution of mountain heights in 10 bins
```{r, echo = F}
options(warn = -1)
graph <- ggplot(hm, aes(x = Height..m.)) + geom_histogram(bins = 10) + labs(x = 'Mountain height in meters', y = 'Count', title = 'Distribution of mountain heights')
#ggsave(plot = graph, filename="../figures/MountainHeights.png")
graph
```

- The distribution is right skewed (median is to the left of the mean), meaning we have a small number of mountains that fall to the right of the mean and a large number that fall to the left of the mean.  

## 2. Grouping the mountain heights into 5 groups based on their height
```{r, echo = F}
graph <- ggplot(hm, aes(x = HeightCategory)) + geom_histogram(bins = 5, stat = 'count') + labs(x = 'Height category', y = 'Count', title = 'Height count when height is categorized into 5 groups')
#ggsave(plot = graph, filename="../figures/MountainHeights5groups.png")
graph
```

- We decided to divide the mountains into 5 categories based on their height to aid us in further analysis of the dataset
- Based on this initial grouping we see the most mountains fall into the first category

## 3. Let us see who are the parent mountains of our mountains when looking at our 5 groups
```{r, echo = F}
filter(hm, !is.na(hm$Parent.mountain)) %>% ggplot(aes(x = HeightCategory, fill = Parent.mountain)) + geom_histogram(bins = 5, color = 'black') + labs(fill = 'Parent mountain', x = 'Height category', y = 'Count', title = 'Height count based on the mountain\'s parent mountain\n\n') -> graph
#ggsave(plot = graph, filename="../figures/MountainHeightsByParent.png")
graph
```

- We wanted the see if we could spot any link between the mountain's height and its parent mountain
- We can see that the highest mountains mostly have Mount Everest as its parent mountain

## 4. Is there any link between the mountain height and the number of failed attempts ?
```{r, echo = F}
graph <- ggplot(hm, aes(x = hm$Height..m., y = hm$Failed.attempts.bef..2004)) + geom_point() + labs(x = 'Mountain height in meters', y = 'Number of failed ascent attempts', title = 'Number of failed ascent attempts in connection with the mountain height')
#ggsave(plot = graph, filename="../figures/FailedAscentAttemptsPoints.png")
graph
```

- Due to small number of data points it is hard to draw any conclusions but a sort of exponential growth is noticable

We can try to visualize the pattern using stat_smooth

```{r, echo = F}
graph + geom_point() + stat_smooth(method = 'loess') -> graph2
#ggsave(plot = graph2, filename="../figures/FailedAscentAttemptsPointsLoess.png")
graph2
```

Or we could furhter try to improve our visualization by using different colors for each height category.
```{r, echo = F}
graph <- ggplot(hm, aes(x = hm$Height..m., y = hm$Failed.attempts.bef..2004, color = HeightCategory)) + geom_point() + geom_smooth(aes(x = hm$Height..m., y = hm$Failed.attempts.bef..2004), method = 'loess', linetype = 4, color = "grey", se = F, inherit.aes = F) + labs(x = 'Mountain height in meters', y = 'Number of failed ascent attempts', title = 'Number of failed ascent attempts in connection with the mountain height')
#ggsave(plot = graph, filename="../figures/FailedAscentAttemptsPointsLoessColor.png")
graph
```


## 5. Let us again see the number of failed attempts for each height category using histogram
```{r, echo = F}
options(warn = -1)
hm %>% group_by(HeightCategory) -> groupedByHC
summarise(groupedByHC, countOfFailures = sum(Failed.attempts.bef..2004, na.rm = T)) -> groupedByHC
graph <- ggplot(groupedByHC, aes(groupedByHC$HeightCategory, groupedByHC$countOfFailures)) + geom_col() + labs(x = 'Height category', y = 'Failed attempts', title = 'Number of failed attempts for each height category')
#ggsave(plot = graph, filename="../figures/FailedAscentAttempts.png")
graph
```

- Using histogram we can see that the third category of mountains was the most 'trialed and errored'

## 6. Let's investigate in which years were the mountain hikers most active
```{r, echo = F}

filter(hm, !is.na(hm$FirstClimbCategory)) %>% ggplot(aes(FirstClimbCategory)) + geom_histogram(bins = 5, stat = 'count', na.rm = T) + labs(x = 'Height category', y = 'Count', title = 'Count of ascents when categorized into 5 groups by the year of ascent') -> graph
#ggsave(plot = graph, filename="../figures/YearOfAscents.png")
graph
```

- We see that according to this dataset most ascents fall in the category that is between years 1946 and 1962.


